---
title: "NSW Road Crash Severity Prediction: Project Plan and EDA"
author: "Group 7, Workshop 12"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    code_folding: hide
---
# Introduction

```{r library, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(readxl)
library(forcats)
library(scales)
library(knitr)
```


```{r read merged dataset, message=FALSE, warning=FALSE}
df_merged <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
```

# 1 Framing & Dataset Selection

## 1.1 Classification Task

Based on environmental conditions and traffic unit characteristics, predict the 
severity of a road crash as a multi-class classification problem: Fatal, 
Injury, or Non-casualty.

                                                      
## 1.2 Research Question

Given environmental conditions (weather, natural lighting, road surface, 
speed limit) and traffic unit characteristics (number of traffic units involved, 
type of traffic units, road classification), can we predict the severity 
of a road crash in NSW?                                                        
  
## 1.3 Motivation                                                

This research question has direct real-world relevance, NSW recorded 340 road 
fatalities in 2024, with over two-thirds occurring in regional areas 
(Transport for NSW, 2025), highlighting the need for data-driven approaches to 
understanding crash severity. Our study affects the following groups:
  
**NSW Traffic Management Authorities**: By identifying key combinations of 
environmental and traffic characteristics that contribute to severe crashes, 
authorities can make targeted decisions such as adjusting speed limits, 
improving road design, and increasing enforcement during high-risk conditions.   

**Emergency Response Teams**: In the early stages of a crash, predicting severity 
based on environmental conditions can help allocate ambulance and emergency 
resources more quickly and effectively.

**Policymakers and Insurance Institutions**: Understanding which traffic unit types 
(e.g. heavy vehicles) pose higher risks under specific environmental conditions 
can inform more precise road safety policies and insurance pricing models.

The dataset contains 92,189 crash records across NSW from 2020 to 2024, covering 
variables such as weather, natural lighting, road surface condition, speed 
limit, number of traffic units involved, and traffic unit types: fully 
supporting the environmental and traffic unit characteristics required by our 
research question.

A predictive model for crash severity enables data-driven prioritization, this 
can help authorities allocate resources design intervention strategies, and form 
policy based on evidence rather than intuition. 

# 2 Consider & Gather Data 

## 2.1 Data Source 

The data used in this study is sourced from the Australian Government open data platform

**NSW Crash Data Dataset:** 
https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data 

This dataset is published by Transport for NSW and provides detailed information on road crashes in NSW, including environmental conditions, road characteristics, 
vehicle types, and injury outcomes (Transport for NSW, 2024). 

For this study, we utilized the Crash Dataset (NSW Road Crash Data - 2020-2024 
- CRASH.xlsx) and Traffic Unit Dataset (NSW Road Crash Data - 2020-2024 
- TRAFFIC UNIT.xlsx ) for 2020-2024.The crash dataset contains environmental and 
road- related variables (e.g. weather, road surface, and speed limit ) 

- The crash dataset contains: 
  - Crash Details (No. Killed, No. Seriously Injured, Severity of Crash) 
  - time-related features (Day, Month, etc) 
  - location attributes (Direction, latitude, longitude, Distance, etc) 
  - environmental conditions (Weather, Natural Lighting, Speed Limit, etc.) 
  
- The traffic unit dataset contains vehicle-level variables: 
  - TU type group (states vehicle type) 
  - Manoeuvre (states vehicle movement at time of crash)
  - TU role in first impact (states role of vehicle e.g. key traffic unit) 
  - TU controlled flag (Whether classified as controlled or uncontrolled)
  
```{r}
cat("Merged:", nrow(df_merged), "rows,", ncol(df_merged), "cols\n")
colSums(is.na(df_merged))
```
  
## 2.2 Dataset Overview

The crash dataset contains 92,189 crash records, and the traffic dataset 
contains 170,962 records across NSW from 2020 to 2024. The dataset formed by 
merging both of these on Crash ID contains 170,962 records. The merged dataset 
is recorded at a vehicle level. This allows vehicle-specific characteristics to 
be incorporated into the crash severity increasing the richness of our dataset.

```{r na visualisation}
na_counts <- colSums(is.na(df_merged))
na_df <- data.frame(
  variable = names(na_counts),
  missing = na_counts
) |> subset(missing > 0)

ggplot(na_df, aes(x = reorder(variable, -missing), y = missing)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(title = "Missing Values by Column", x = "Column", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2.3 Dataset Justification

The dataset meets several key requirements outlined in the assignment.

**Large Dataset(>10,000 observations)**

The merged dataset has 170,962 records, providing sufficient scale for robust 
statistical analysis and model training. The large size enables reliable 
estimation of relationships between our target variable (crash severity) and its 
predictors while also supporting validation techniques such as test-train split.

**Integrated Dataset (Multiple sources combined)**

The dataset is formed by merging crash-level and traffic unit datasets using a 
common Crash ID. This integration combines environmental, spatial, and vehicle-
level information. The process improves the richness of features available for 
analysis and enables more comprehensive modelling of crash severity.

**Messy Dataset (Real-world data challenge)**

As a real-world administrative dataset, it contains missing values and potential 
inconsistencies across variables such as weather, lighting, and vehicle 
characteristics. Additionally, the process of merging the crash dataset (92,189 
records) with traffic unit dataset (170,962 records) exposes existing missing
values, as not all variables are available at both the crash and vehicle levels. 
When merged using Crash ID, crash-level attributes are duplicated across 
multiple vehicle records, while some vehicle-specific fields may not have 
corresponding values for all observations.

This results in a dataset with increased sparsity and inconsistencies across 
columns, requiring preprocessing steps such as handling missing data, managing 
duplicated information, and cleaning categorical variables. As shown in the 
above figure, columns such as Primary hazardous feature, Primary temporary 
feature, and object hit 2 have the highest missingness(>95%) while Route no. and 
Other TU type have moderate missingness

**Multi-class Classification Problem**

The target variable, crash severity, is a multi-class outcome derived from 
Degree of crash column in the crash dataset, consisting of  categories such as 
Fatal, Injury, and Non-casualty. These categories are supported by injury-related 
variables in the crash dataset (e.g., number killed and 
number seriously injured). This introduces challenges such as class imbalance, 
which must be considered when evaluating model performance.

**2.4 Summary**

Overall, the dataset is well-suited for this study as it is large, integrated, 
messy, and supports a multi-classification problem. Its combination of 
environmental, temporal, spatial, and vehicle-level features provides a strong 
foundation for modelling and predicting crash severity.

# 3 Exploratory Data Analysis (EDA)
## 3.1 Univariate Analysis
### 3.1.1 Distribution
```{r barpolot of degree of crash}
ggplot(df_merged, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")
```
```{r barplot of weather}
ggplot(df_merged, aes(x = fct_infreq(Weather))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Weather Conditions",
       x = "Weather", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```
```{r barplot of natrual lighting}
ggplot(df_merged, aes(x = fct_infreq(`Natural lighting`))) +
  geom_bar(fill = "orange") +
  labs(title = "Distribution of Natural Lighting", x = "Lighting", y = "Count") +
  theme_minimal()
```
```{r bar plot of speed limit}
df_merged$`Speed limit` <- factor(df_merged$`Speed limit`,
                              levels = c("10 km/h", "20 km/h", "30 km/h", "40 km/h",
                                         "50 km/h", "60 km/h", "70 km/h", "80 km/h",
                                         "90 km/h", "100 km/h", "110 km/h", "Unknown"))

ggplot(df_merged, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of Surface condition}
ggplot(df_merged, aes(x = `Surface condition`)) + 
  geom_bar(fill = "darkgreen") +
  labs(title = "Distribution of Surface condition", x = "Surface condition", y = "Count") +
  theme_minimal()+
  scale_y_continuous(labels = scales::comma)
```
```{r barplot of Urbanisation}
ggplot(df_merged, aes(x = fct_infreq(Urbanisation))) + 
  geom_bar(fill = "purple") + 
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r bar plot of TU type group}
ggplot(df_merged, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of Traffic Unit Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of Manoeuvre}
ggplot(df_merged, aes(x = fct_infreq(Manoeuvre))) +
    geom_bar(fill = "coral") +
    labs(title = "Distribution of Manoeuvre", x = "Manoeuvre", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of TU role in first impact}
ggplot(df_merged, aes(x = `TU role in first impact`)) +
    geom_bar(fill = "gold") +
    labs(title = "Distribution of TU Role in First Impact", x = "Role", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of TU group}
df_tu <- df_merged %>%
    mutate(`TU group` = ifelse(`No. of traffic units involved` >= 6, "6+",
                                as.character(`No. of traffic units involved`)))                                                                            
df_tu$`TU group` <- factor(df_tu$`TU group`, levels = c("1", "2", "3", "4", "5", "6+"))                                                                  
ggplot(df_tu, aes(x = `TU group`)) +                                                                                                                     
geom_bar(fill = "slateblue") +                          
labs(title = "Distribution of No. of Traffic Units Involved", x = "No. of Traffic Units", y = "Count") + theme_minimal() 
```

The data shows significant class imbalance: Injury crashes are most frequent (114,355), followed by Non-casualty (54,191), with Fatal crashes being the minority (2,416). Most crashes occur in fine weather, daylight, and dry road conditions, suggesting traffic volume has more influence than weather on crash frequency. The 50-60 km/h speed zones account for the most crashes, with Sydney metro being the dominant urbanisation category. Car/car derivatives are the most common vehicle type, "Proceeding in lane" is the most frequent manoeuvre (~100,000), with all other manoeuvres being substantially less frequent. Key traffic units account for the largest share of roles in the first impact, and 2-vehicle crashes are the most common collision type.

### 3.1.2 Missing Values
```{r na check}
colSums(is.na(df_merged))
```

Core modelling variables (Weather, Speed limit, Natural lighting, Degree of crash) have no missing values. Primary hazardous feature (168,803), Primary temporary feature (168,406), and Object hit 2 (165,996) have very high missingness due to their conditional nature and will be dropped during preprocessing.

### 3.1.3 Outliers for Key Variables
```{r summary of TU group}
summary(df_merged$`No. of traffic units involved`)
```

```{r}
ggplot(df_merged, aes(y = `No. of traffic units involved`)) +
  geom_boxplot(fill = "steelblue", outlier.colour = "red", outlier.size = 1) +
  scale_y_log10() +
  labs(title = "Boxplot of No. of Traffic Units Involved (Log Scale)",
       y = "No. of Traffic Units (log scale)") +
  theme_minimal()
```

The number of traffic units has a median of 2 and mean of 2.149, with IQR concentrated between 1 and 2. The maximum value of 26 is an extreme outlier and may be capped during preprocessing.

## 3.2 Bivariate Analysis
### 3.2.1 Features vs Outcomes
```{r weather vs degree of crash}
ggplot(df_merged, aes(x = Weather, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Weather", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r speed limit vs degree of crash}
ggplot(df_merged, aes(x = `Speed limit`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Speed Limit", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r natural lighting vs degree of crash}
ggplot(df_merged, aes(x = `Natural lighting`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Natural Lighting", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r TU type group vs degree of crash}
ggplot(df_merged, aes(x = `TU type group`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Type", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r manoeuvre vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(Manoeuvre)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Manoeuvre", x = "Manoeuvre", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```
```{r Urbanisation vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(Urbanisation)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Urbanisation", x = "Urbanisation", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r Surface condition vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(`Surface condition`)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Surface condition", x = "Surface condition", y = "Proportion") +
  theme_minimal()
```
```{r TU controlled flag vs degree of crash}
ggplot(df_merged, aes(x = `TU controlled flag`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Controlled Flag", x = "Controlled", y = "Proportion") +
  theme_minimal()
```

Speed limit shows the strongest relationship with crash severity, with higher speed zones (100-110 km/h) having noticeably greater Fatal proportions. Pedestrians and motorcycles show elevated Injury proportions, while uncontrolled traffic units have higher Non-casualty proportions. Country non-urban areas show slightly higher Fatal proportions compared to urban areas, consistent with higher rural speed limits.

### 3.2.2 Features vs Features
```{r weather vs natural lighting}
df_merged %>%
  group_by(Weather, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Natural Lighting", fill = "Count") +
  theme_minimal()
```

```{r speed limit vs urbanisation}
df_merged %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Urbanisation, y = `Speed limit`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Speed Limit vs Urbanisation", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r Weather vs Surface condition}
df_merged %>%
  group_by(Weather, `Surface condition`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Surface condition` != "Unknown") %>%
  ggplot(aes(x = `Surface condition`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Surface Condition", fill = "Count") +
  theme_minimal()
```

```{r TU type group vs Manoeuvre}
df_merged %>%
  group_by(`TU type group`, Manoeuvre) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Manoeuvre, y = `TU type group`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "TU Type vs Manoeuvre", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```

```{r Urbanisation vs Natural lighting}
df_merged %>%
  group_by(Urbanisation, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(`Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Urbanisation, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Urbanisation vs Natural Lighting", fill = "Count") +
  theme_minimal()
```

Fine weather strongly aligns with dry surfaces while rain almost exclusively pairs with wet conditions, confirming dataset consistency. Speed limits of 50-60 km/h dominate Sydney metro areas while 100-110 km/h zones concentrate in Country non-urban regions. Crashes predominantly occur in daylight across all urbanisation categories, with Sydney metro having the highest overall crash volume.

## 3.3 Data Visualisation: Additional Graphs/Tables with Insights
To further support the exploratory analysis, additional visualizations and summary tables were produced to reveal temporal, spatial, and contextual patterns in crash severity. These displays go beyond simple univariate and bivariate summaries by highlighting high-risk combinations of conditions and identifying patterns that may influence feature selection and model choice.

### 3.3.1 Heatmap-day-time
```{r eda-helpers, message=FALSE, warning=FALSE}
library(knitr)

# Ensure ordered factors for time-related variables
day_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
df_merged$`Day of week of crash` <- factor(df_merged$`Day of week of crash`, levels = day_levels)

# If Two-hour intervals is character, keep original order as it appears
df_merged$`Two-hour intervals` <- factor(df_merged$`Two-hour intervals`,
                                         levels = unique(df_merged$`Two-hour intervals`))

# Order reporting year if needed
df_merged$`Reporting year` <- factor(df_merged$`Reporting year`,
                                     levels = sort(unique(df_merged$`Reporting year`)))
```
```{r heatmap-day-time}
df_merged %>%
  group_by(`Day of week of crash`, `Two-hour intervals`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = `Two-hour intervals`, y = `Day of week of crash`, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Crash Frequency by Day of Week and Time Interval",
    x = "Two-hour Interval",
    y = "Day of Week",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The heatmap shows that crash frequency is concentrated in specific time periods rather than being evenly distributed across the week. In particular, daytime and peak commuting periods tend to have the highest crash counts, suggesting that traffic exposure and congestion may play a major role in crash occurrence. This pattern indicates that time-related variables are likely to be useful predictors in the classification task.

### 3.3.2 Severity-by-year
```{r severity-by-year}
df_merged %>%
  group_by(`Reporting year`, `Degree of crash`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = `Reporting year`, y = count, fill = `Degree of crash`)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Crash Severity by Reporting Year",
    x = "Reporting Year",
    y = "Proportion",
    fill = "Crash Severity"
  ) +
  theme_minimal()
```
The proportional distribution of crash severity appears relatively stable across reporting years, with Injury crashes consistently dominating the dataset and Fatal crashes remaining the minority class. This suggests that class imbalance is a persistent characteristic of the data rather than a one-off issue in a single year, reinforcing the need for imbalance-aware modelling and evaluation.

### 3.3.3 Fatal-rate-speed-urban
```{r fatal-rate-speed-urban}
fatal_heatmap <- df_merged %>%
  filter(!is.na(`Speed limit`), !is.na(Urbanisation), !is.na(`Degree of crash`)) %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(
    total = n(),
    fatal_rate = mean(`Degree of crash` == "Fatal"),
    .groups = "drop"
  ) %>%
  filter(total >= 100)  # avoid unstable rates from tiny groups

ggplot(fatal_heatmap, aes(x = Urbanisation, y = `Speed limit`, fill = fatal_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkred",
                      labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Fatal Crash Proportion by Speed Limit and Urbanisation",
    x = "Urbanisation",
    y = "Speed Limit",
    fill = "Fatal Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The heatmap suggests that the proportion of Fatal crashes is higher in high-speed areas, especially in less urbanised regions. This supports the earlier bivariate findings and indicates that speed environment and spatial context jointly influence crash severity. These interaction patterns justify the use of flexible models such as decision trees, random forests, and boosting methods.

### 3.3.4 High-Risk Crash Scenarios
```{r high-risk-scenarios-table}
risk_table <- df_merged %>%
  filter(
    !is.na(`Speed limit`),
    !is.na(`Natural lighting`),
    !is.na(Urbanisation),
    !is.na(`Degree of crash`)
  ) %>%
  group_by(`Speed limit`, `Natural lighting`, Urbanisation) %>%
  summarise(
    total_crashes = n(),
    fatal_prop = mean(`Degree of crash` == "Fatal"),
    injury_prop = mean(`Degree of crash` == "Injury"),
    non_casualty_prop = mean(`Degree of crash` == "Non-casualty (towaway)"),
    .groups = "drop"
  ) %>%
  filter(total_crashes >= 200) %>%   # keep only meaningful groups
  arrange(desc(fatal_prop), desc(total_crashes)) %>%   # sort by fatal proportion descending
  slice_head(n = 10) %>%
  mutate(
    fatal_prop = scales::percent(fatal_prop, accuracy = 0.1),
    injury_prop = scales::percent(injury_prop, accuracy = 0.1),
    non_casualty_prop = scales::percent(non_casualty_prop, accuracy = 0.1)
  ) %>%
  rename(
    `Speed Limit` = `Speed limit`,
    `Natural Lighting` = `Natural lighting`,
    `Total Crashes` = total_crashes,
    `Fatal Proportion` = fatal_prop,
    `Injury Proportion` = injury_prop,
    `Non-casualty Proportion` = non_casualty_prop
  )

knitr::kable(
  risk_table,
  caption = "Table: High-risk crash scenarios ranked by fatal crash proportion"
)
```
The table highlights combinations of environmental and contextual conditions associated with relatively high fatal proportions. Rather than relying on a single variable in isolation, these results suggest that crash severity is shaped by combinations of speed environment, lighting conditions, and urban context. This motivates the inclusion of interaction-capable models in the modelling phase.

### 3.3.5 Top 10 LGA by crash count
```{r top-lga}
top_lga <- df_merged %>%
  filter(!is.na(LGA)) %>%
  count(LGA, sort = TRUE) %>%
  slice_head(n = 10)

ggplot(top_lga, aes(x = reorder(LGA, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 LGAs by Crash Count",
    x = "LGA",
    y = "Number of Crashes"
  ) +
  theme_minimal()
```
Crash counts are unevenly distributed across LGAs, indicating spatial heterogeneity in crash occurrence. This suggests that location-related variables such as LGA and urbanisation may add predictive value, although high-cardinality location variables may require grouping or regularisation during preprocessing.

### 3.3.6 Fatal-rate-tu-type
```{r fatal-rate-tu-type}
tu_fatal <- df_merged %>%
  filter(!is.na(`TU type group`), !is.na(`Degree of crash`)) %>%
  group_by(`TU type group`) %>%
  summarise(
    total = n(),
    fatal_rate = mean(`Degree of crash` == "Fatal"),
    .groups = "drop"
  ) %>%
  filter(total >= 100) %>%
  arrange(desc(fatal_rate))

ggplot(tu_fatal, aes(x = reorder(`TU type group`, fatal_rate), y = fatal_rate)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Fatal Crash Proportion by Traffic Unit Type",
    x = "Traffic Unit Type",
    y = "Fatal Proportion"
  ) +
  theme_minimal()
```
Fatal crash proportions vary noticeably across traffic unit types, suggesting that not all road users face the same severity risk. This reinforces the importance of traffic-unit-level features in the integrated dataset and supports the decision to retain traffic-unit characteristics for modelling.


# 4 Project Planning
## 4.1 Preprocessing Plan for handling your dataset’s specific requirements to make it clean and ready for modelling.

To transform the raw NSW Road Crash Data into a clean, model-ready dataset, we propose the following preprocessing pipeline. All data-dependent steps (imputation, encoding, scaling, resampling) will be performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

### Step 1: Data Integration

The dataset consists of two linked tables — **CRASH** (92,189 rows × 50 columns) and **TRAFFIC UNIT** (170,962 rows × 10 columns) — connected via `Crash ID`. Since this is a one-to-many relationship, we will **aggregate** the TRAFFIC UNIT table to crash level before joining:

-   Binary flags for vehicle type presence: `has_heavy_vehicle`, `has_motorcycle`, `has_pedestrian`, `has_bicycle`
-   Count of distinct TU types per crash
-   Binary flag for whether any object was hit

The aggregated table will be **left-joined** to CRASH on `Crash ID`, producing a single analysis-ready table.

### Step 2: Feature Selection & Removal

| Category | Columns | Reason |
|----|----|----|
| Identifiers | `Crash ID` | Not predictive |
| Target leakage | `Degree of crash - detailed`, `No. killed`, `No. seriously injured`, `No. moderately injured`, `No. minor-other injured` | Direct consequences of severity — would leak the target |
| Redundant location | `Street of crash`, `Identifying feature`, `Town`, `Route no.`, `Street type`, `Direction` | High cardinality; location already captured by `LGA`, `Lat/Lon`, `Urbanisation` |
| Near-zero variance | `Primary temporary feature` (98.5% NA), `Primary hazardous feature` (98.0% NA), `DCA supplement` (86.8% NA) | Too sparse to be informative |

Retained features cover **environmental conditions** (weather, lighting, road surface, speed limit), **road characteristics** (alignment, type of location), **temporal patterns** (month, day of week, time interval), **spatial context** (urbanisation, LGA), and **traffic unit characteristics** (aggregated from TU table).

### Step 3: Missing Value Handling

| Variable | Missing % | Strategy |
|----|----|----|
| `Primary permanent feature` | 73.0% | Convert to binary indicator `has_permanent_feature`; drop original |
| `Other TU type` | 30.4% | Recode `NA` → `"Single vehicle"` (missing implies single-vehicle crash) |
| Remaining categorical | \<1% | Treat `NA` as an explicit `"Unknown"` level |
| Remaining numeric | \<1% | Median imputation (within training fold) |

For categorical variables, treating missing as its own category is justified when the missingness pattern itself is informative (Little, 1988). We will verify this using `naniar::mcar_test()` and visualise with `visdat::vis_miss()`.

### Step 4: Feature Engineering

Derived features to enhance predictive power:

-   **Temporal**: `is_weekend` (Sat/Sun), `is_rush_hour` (7–9am / 4–6pm)
-   **Environmental**: `is_night` (Darkness/Twilight), `is_wet` (non-Dry surface)
-   **Speed bands**: Group `Speed limit` into `≤50`, `60–80`, `90–100`, `110+`, `Other`
-   **Multi-vehicle**: Binary flag from `No. of traffic units involved > 1`

### Step 5: Outlier Detection

-   **Latitude/Longitude**: Remove records outside NSW bounds (lat: −37.5 to −28.0, lon: 140.0 to 154.0)
-   **Distance**: Inspect for extreme values; winsorise at the 99th percentile if needed

### Step 6: Encoding Categorical Variables

Encoding strategy is model-dependent:

-   **Tree-based models** (Random Forest, XGBoost): Use R `factor` type directly — no encoding needed
-   **Distance/linear models** (kNN, SVM, Logistic Regression, LDA): One-hot encoding for nominal variables; ordinal encoding for ordered variables (e.g., speed bands)
-   **High-cardinality**: Pool `LGA` levels with \<1% of observations into `"Other"` to reduce dimensionality

### Step 7: Feature Scaling

-   **kNN and SVM** are distance-based and sensitive to feature scale → apply **z-score standardisation** to all numeric features within each training fold
-   **Tree-based models** are scale-invariant → no scaling needed

### Step 8: Class Imbalance Handling

The target `Degree of crash` is highly imbalanced (Fatal: 1.5%, Injury: 66.4%, Non-casualty: 32.1%). Resampling will be applied **within the training fold only** to prevent leakage (Chawla et al., 2002):

1.  **SMOTE** (`themis::step_smote`): Synthesise minority-class samples, especially for Fatal
2.  **Down-sampling**: Undersample the majority class as a simpler alternative
3.  Both strategies will be compared during model tuning to select the best approach

### Step 9: Train/Test Split & Cross-Validation

-   **Holdout**: 80/20 stratified split by `Degree of crash`
-   **Nested CV**: Outer 5-fold for generalisation error; inner 5-fold for hyperparameter tuning
-   All preprocessing steps will be embedded in a `tidymodels::recipe` pipeline so that transformations are fitted on training data only within each fold

### Pipeline Summary

```         
Raw Data (CRASH + TRAFFIC UNIT)
  │
  ├─ 1. Aggregate TRAFFIC UNIT → crash-level features
  ├─ 2. Left join on Crash ID
  ├─ 3. Remove identifiers, leakage vars, near-zero-variance cols
  ├─ 4. Feature engineering (binary flags, speed bands, temporal)
  ├─ 5. Outlier removal (geospatial bounds)
  ├─ 6. Train/Test split (80/20, stratified)
  │
  └─ Within each CV fold (training only):
       ├─ 7. Missing value handling
       ├─ 8. Categorical encoding (one-hot / factor)
       ├─ 9. Feature scaling (for kNN/SVM/LR)
       └─ 10. SMOTE / down-sampling for class imbalance
```

## 4.2 Model Plan: based on EDA, propose at least 5 classification models* suited to the data/problem with justification.

Based on the findings from EDA, we propose six classification models to predict **Degree of Crash** (a three-class outcome: Fatal, Injury, and Non-casualty). The EDA revealed that the target variable is extremely imbalanced—Fatal crashes account for a small proportion of the 92,189 observations—and that all predictor variables are categorical (e.g., `Weather`, `Natural lighting`, `Surface condition`, `Speed limit`, `Urbanisation`, `TU type group`). These characteristics directly inform the choice of models below. 

All models will be implemented in R using the `caret` package for a unified training and cross-validation interface.

### 1. Multinomial Logistic Regression (Baseline)
- **Justification:** Proposed as our interpretable baseline model. It produces interpretable coefficients (log-odds) for each predictor category, which are meaningful for road safety stakeholders (e.g., quantifying how much driving at night increases the log-odds of a fatal crash relative to daylight). All subsequent models will be benchmarked against this baseline.
- **Implementation:** Categorical predictors will be encoded as dummy variables. We will use `multinom()` from the `nnet` package, utilising the `class_weight` argument to account for class imbalance.

### 2. Decision Tree Classifier (CART)
- **Justification:** Chosen for its full interpretability and ability to capture non-linear interactions between categorical features (e.g., capturing the joint rule "high speed limit AND rural area $\rightarrow$ higher fatal risk" in a single branch).
- **Implementation:** Implemented using the `rpart` package. Tree depth will be tuned via cross-validation to prevent overfitting, and the resulting tree will be visualised using `rpart.plot` to communicate high-risk conditions to non-technical audiences.

### 3. Random Forest Classifier
- **Justification:** Extends the decision tree by training an ensemble of trees on bootstrap samples, significantly reducing variance. It provides variable importance scores to identify which predictors most strongly impact crash severity. 
- **Implementation:** With >90k observations, it is computationally feasible using the `randomForest` package. Class imbalance will be handled via the `classwt` argument in `randomForest()`.

### 4. Gradient Boosting Machine (XGBoost)
- **Justification:** Proposed as our primary high-performance model. By sequentially correcting misclassifications from prior trees, it is particularly effective at learning from the rare minority class (Fatal crashes) by exploiting subtle, high-value signals in the features.
- **Implementation:** Implemented using the `xgboost` package. Hyperparameters (`nrounds`, `max_depth`, `eta`) will be tuned via `caret::trainControl`. The `scale_pos_weight` parameter will be adjusted to address the class imbalance.

### 5. K-Nearest Neighbours (KNN)
- **Justification:** Included as a non-parametric, instance-based comparison model to test whether crash severity is locally consistent in feature space (i.e., whether crashes occurring under similar conditions tend to share the same outcome).
- **Implementation:** Implemented using the `class` package. Since all features are categorical, the Gower distance metric (via the `gower` package) will be used. The optimal value of $k$ will be selected via cross-validation.

### 6. Naive Bayes
- **Justification:** A natural fit for datasets where all features are categorical. It serves as a strong probabilistic benchmark: if more complex models cannot substantially outperform it, this suggests that the individual distributions of features (e.g., Weather, Speed limit) already carry most of the predictive information.
- **Implementation:** Implemented using the `e1071` package. Under the conditional independence assumption, it is extremely fast to train and provides probability estimates for each severity class.

## 4.3 Evaluation Plan: propose and justify performance metrics for final model 
evaluation.

Standard accuracy is an inadequate evaluation metric for this problem due to the class imbalance in Degree of Crash. At the vehicle level, Fatal observations account for only 1.4% of the 170,962 records, with Injury at 66.9% and Non-casualty at 31.7%. A classifier that predicts Injury for every observation would achieve high accuracy simply because Injury is the most frequent class, while completely failing to identify any Fatal crash. The metrics below are selected to reflect both overall classifier performance and the disproportionate importance of correctly identifying fatal outcomes.

###Primary Metric: Macro-Averaged F1 Score
The macro-averaged F1 score computes the F1 score (harmonic mean of precision and recall) for each of the three classes independently — Fatal, Injury, and Non-casualty — and then averages them with equal weight, regardless of class frequency. This ensures that performance on the rare Fatal class receives the same weight as the dominant Injury class, directly penalising models that achieve high overall accuracy by ignoring fatal crashes. Macro F1 is the primary metric for ranking and selecting the final model.

###Primary Constraint: Fatal Class Recall
Given the real-world stakes of missing a fatal crash, recall for the Fatal class is reported separately as a hard constraint alongside macro F1. Recall measures the proportion of actual Fatal crashes that the model correctly identifies. A model with high macro F1 but low Fatal recall will be deprioritised in model selection, as failing to flag potentially fatal road conditions is the most consequential error in this problem context.

###Secondary Metric: Weighted F1 Score
The weighted F1 score averages F1 per class weighted by each class's support (number of observations). This provides a realistic picture of overall classifier performance on the true class distribution and complements macro F1 by quantifying the trade-off between performance on rare versus frequent classes.

###Secondary Metric: Cohen's Kappa (κ)
Cohen's Kappa adjusts accuracy for the agreement expected by chance, given the class distribution. In an imbalanced three-class problem, even a naive classifier that always predicts the most frequent class has a non-trivial chance of being correct. Kappa values near 0 indicate performance no better than random guessing. Kappa is computed directly by caret's confusionMatrix() function and provides a clear sanity check on all models.

###Secondary Metric: Confusion Matrix
A confusion matrix will be produced for each model using caret's confusionMatrix() function and visualised as a heatmap using ggplot2. The 3×3 matrix reveals which severity classes are most frequently misclassified and in which direction — for example, whether Fatal crashes are being predicted as Injury, which is the most dangerous type of error in this context.

###Validation Strategy
Cross-validation folds are constructed at the crash level (grouped by Crash ID) to prevent data leakage from multiple vehicles belonging to the same crash appearing in both training and validation folds. Stratified grouped 5-fold cross-validation will be applied to the training data using caret's groupKFold() with Crash ID as the grouping variable, preserving class proportions across folds. A temporal split will be used: 2020–2023 data for training and 2024 data as the held-out test set, reflecting real-world deployment conditions where the model is used to predict future crashes. The final held-out test set is evaluated once only, after model selection is complete via cross-validation, to obtain unbiased estimates of generalisation performance.

## 4.4 Timeline and Project Milestones

The project timeline has been structured in alignment with the assignment 
deadlines to ensure clear and systematic progression of tasks from understanding 
the dataset up-to final model development and reporting. 

**1-7 April: Problem Framing and Data Acquisition**

Different datasets were evaluated and a final dataset was selected, along with 
finalization of the research question. The crash and Traffic Unit datasets (2020
-2024) were sourced from the NSW Government open data platform. Initial 
inspection was carried out to understand the dataset structure, key variables, 
and differences. 

**7-10 April: Data Integration and Initial Data Preparation**

The crash and traffic unit datasets were merged using Crash ID, resulting in a 
unified dataset containing 170,962 records at the vehicle level. Initial data 
preparation involved examining missing values, identifying inconsistencies 
introduced via the merging, and understanding different variable types. Full 
preprocessing and feature engineering will be conducted in the modelling phase. 

**10-14 April: Exploratory Data Analysis (EDA)**

Univariate and bivariate analyses were performed to explore the distribution and 
relationship between variables. This includes examining environmental factors 
(weather, lighting), temporal variables(time of day, day of week), and vehicle-
level characteristics (vehicle type, manoeuvre). Insights from EDA were used to 
inform feature selection and guide model planning.  

**14-17 April: Feature Planning and Model Design**

Based on EDA findings, relevant features were identified and transformations 
were planned, including the plan to encode categorical variables and aggregate 
vehicle-level information. A set of classification models were also selected for 
implementation in the next phase.

**17-19 April: Plan & EDA Report Finalization**

All findings were compiled into a structured R Markdown report and submitted by 
the deadline.

**6–10 May: Data Preprocessing and Baseline Modelling**

Preprocessing steps such as handling missing values, encoding categorical variables, and feature engineering will be carried out. The baseline model (Multinomial Logistic Regression) will also be implemented and evaluated.

**11–15 May: Advanced Model Development and Tuning**

Advanced models will be implemented and refined, including hyperparameter tuning via cross-validation to improve performance.

**16–20 May: Model Evaluation and Report Writing**

Models will be evaluated using appropriate performance metrics, considering class imbalance in crash severity. The final model will be selected and justified. Results, insights, and interpretations will be consolidated into the final report.

**21–24 May: Report Refinement**

Final refinements will be made to ensure clarity, consistency, and adherence to the submission requirements.

**25–31 May: Final Checks and Contribution Review**

Team contributions will be reviewed and documented. A final check of the report will be conducted before submission.

**Backup Plan**

The team acknowledges that unforeseen circumstances such as illness, personal 
emergencies, or other commitments may arise during the project period. 
With this in mind, the following contingency measures have been put in place. 
The workload of members has been structured so that 2 members are on standby at 
any given time, ready to take over tasks if a team member is unable to meet a 
deadline. The remaining members will provide 2 check-ins per week to monitor 
task progress and flag any delays early. If a task is not completed by its 
deadline, it will be immediately handed over to a standby member to minimise 
disruption to the overall timeline. All work will be stored in a shared 
github repository accessible to all team members to ensure continuity.

The team will hold 2 meetings per week during project periods. One every Thursday 
in-person after the workshop, and one every Monday via Zoom to track progress, 
update team members, and provide support. Each member is expected to contribute 
approximately 20% to the overall project across both phases.

# References

Chawla, N. V., Bowyer, K. W., Hall, L. O., & Kegelmeyer, W. P. (2002). SMOTE: Synthetic minority over-sampling technique. *Journal of Artificial Intelligence Research*, *16*, 321–357. https://doi.org/10.1613/jair.953

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning: With applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Little, R. J. A. (1988). A test of missing completely at random for multivariate data with missing values. *Journal of the American Statistical Association*, *83*(404), 1198–1202. https://doi.org/10.1080/01621459.1988.10478722

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. Australian Government Open Data Platform. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety. https://www.transport.nsw.gov.au/system/files/media/documents/2025/crs_road_safety_progress_report_2024.pdf

# Acknowledgement of AI

Group 07 (Workshop 12) declares that all ideas, analysis, and written content 
presented in this report are our own. We did not use generative AI to write our 
code. We used ChatGPT (GPT-5.3, OpenAI, https://chatgpt.com) and Claude Sonnet 4.5 
(Anthropic, https://claude.ai) for code refinement, grammar checking, and 
reviewing our report against the marking rubric.
