---
title: "NSW Road Crash Severity Prediction: Project Plan and EDA"
author: "Group 7, Workshop 12"
output:
  html_document:
    code_folding: hide
---
# Introduction

```{r library, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(readxl)
library(forcats)
library(scales)
library(knitr)
```


```{r read merged dataset, message=FALSE, warning=FALSE}
df_crash <- read_xlsx("nsw_road_crash_data_2020-2024_crash.xlsx")
df_unit <- read_xlsx("nsw_road_crash_data_2020-2024_traffic_unit.xlsx")
df_merged <- left_join(df_crash, df_unit, by = "Crash ID")
write.csv(df_merged, "nsw_road_crash_merged_2020_2024.csv", row.names = FALSE)
```

# 1 Framing & Dataset Selection

## 1.1 Classification Task

Based on environmental conditions and traffic unit characteristics, predict the 
severity of a road crash as a multi-class classification problem: Fatal, 
Injury, or Non-casualty.

                                                      
## 1.2 Research Question

Given environmental conditions (weather, natural lighting, road surface, 
speed limit) and traffic unit characteristics (number of traffic units involved, 
type of traffic units, road classification), can we predict the severity 
of a road crash in NSW?                                                        
  
## 1.3 Motivation                                                

This research question has direct real-world relevance, NSW recorded 340 road 
fatalities in 2024, with over two-thirds occurring in regional areas 
(Transport for NSW, 2025), highlighting the need for data-driven approaches to 
understanding crash severity. Our study affects the following groups:
  
**NSW Traffic Management Authorities**: By identifying key combinations of 
environmental and traffic characteristics that contribute to severe crashes, 
authorities can make targeted decisions such as adjusting speed limits, 
improving road design, and increasing enforcement during high-risk conditions.   

**Emergency Response Teams**: In the early stages of a crash, predicting severity 
based on environmental conditions can help allocate ambulance and emergency 
resources more quickly and effectively.

**Policymakers and Insurance Institutions**: Understanding which traffic unit types 
(e.g. heavy vehicles) pose higher risks under specific environmental conditions 
can inform more precise road safety policies and insurance pricing models.

The dataset contains 92,189 crash records across NSW from 2020 to 2024, covering 
variables such as weather, natural lighting, road surface condition, speed 
limit, number of traffic units involved, and traffic unit types: fully 
supporting the environmental and traffic unit characteristics required by our 
research question.

A predictive model for crash severity enables data-driven prioritization, this 
can help authorities allocate resources design intervention strategies, and form 
policy based on evidence rather than intuition. 

# 2 Consider & Gather Data 

## 2.1 Data Source 

The data used in this study is sourced from the Australian Government open data platform

**NSW Crash Data Dataset:** 
https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data 

This dataset is published by Transport for NSW and provides detailed information on road crashes in NSW, including environmental conditions, road characteristics, 
vehicle types, and injury outcomes (Transport for NSW, 2024). 

For this study, we utilized the Crash Dataset (NSW Road Crash Data - 2020-2024 
- CRASH.xlsx) and Traffic Unit Dataset (NSW Road Crash Data - 2020-2024 
- TRAFFIC UNIT.xlsx ) for 2020-2024. The crash dataset contains environmental and road-related variables (e.g. weather, road surface, and speed limit).

- The crash dataset contains: 
  - Crash Details (No. Killed, No. Seriously Injured, Severity of Crash) 
  - time-related features (Day, Month, etc) 
  - location attributes (Direction, latitude, longitude, Distance, etc) 
  - environmental conditions (Weather, Natural Lighting, Speed Limit, etc.) 
  
- The traffic unit dataset contains vehicle-level variables: 
  - TU type group (states vehicle type) 
  - Manoeuvre (states vehicle movement at time of crash)
  - TU role in first impact (states role of vehicle e.g. key traffic unit) 
  - TU controlled flag (Whether classified as controlled or uncontrolled)
  
```{r, results='hide'}
cat("Merged:", nrow(df_merged), "rows,", ncol(df_merged), "cols\n")
colSums(is.na(df_merged))
```
  
## 2.2 Dataset Overview

The crash dataset contains 92,189 crash records, and the traffic dataset 
contains 170,962 records across NSW from 2020 to 2024. The dataset formed by 
merging both of these on Crash ID contains 170,962 records. The merged dataset 
is recorded at a vehicle level. This allows vehicle-specific characteristics to 
be incorporated into the crash severity increasing the richness of our dataset.

```{r na visualisation, fig.width=7, fig.height=3}
na_counts <- colSums(is.na(df_merged))
na_df <- data.frame(
  variable = names(na_counts),
  missing = na_counts
) |> subset(missing > 0)

ggplot(na_df, aes(x = reorder(variable, -missing), y = missing)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(title = "Missing Values by Column", x = "Column", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2.3 Dataset Justification

The dataset meets the four key requirements outlined in the assignment:

| Requirement | How it is met |
|-------------|---------------|
| **Large (>10,000 obs.)** | The merged dataset has 170,962 records, providing sufficient scale for robust statistical analysis, reliable estimation of predictor–severity relationships, and support for train/test validation. |
| **Integrated (multiple sources)** | Formed by merging crash-level and traffic unit datasets via `Crash ID`, combining environmental, spatial, and vehicle-level information to enable more comprehensive modelling. |
| **Messy (real-world challenges)** | Contains missing values and inconsistencies across weather, lighting, and vehicle variables. Merging introduces sparsity: crash-level attributes are duplicated across vehicle records while some vehicle fields lack values. High-missingness columns (Primary hazardous feature >95%, Primary temporary feature >98%) require dropping; moderate-missingness columns (Route no., Other TU type) require imputation strategies. |
| **Multi-class classification** | The target `Degree of crash` has three classes — Fatal, Injury, and Non-casualty — with significant class imbalance (Fatal: ~1.4%), requiring imbalance-aware modelling and evaluation. |

**2.4 Summary**

Overall, the dataset is well-suited for this study as it is large, integrated, messy, and supports a multi-class classification problem. Its combination of environmental, temporal, spatial, and vehicle-level features provides a strong foundation for modelling and predicting crash severity.

# 3 Exploratory Data Analysis (EDA)
## 3.1 Univariate Analysis
### 3.1.1 Distribution
```{r 311-row1, out.width='25%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
ggplot(df_merged, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")

ggplot(df_merged, aes(x = fct_infreq(Weather))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Weather Conditions", x = "Weather", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)

ggplot(df_merged, aes(x = fct_infreq(`Natural lighting`))) +
  geom_bar(fill = "orange") +
  labs(title = "Distribution of Natural Lighting", x = "Lighting", y = "Count") +
  theme_minimal()

df_merged$`Speed limit` <- factor(df_merged$`Speed limit`,
                              levels = c("10 km/h", "20 km/h", "30 km/h", "40 km/h",
                                         "50 km/h", "60 km/h", "70 km/h", "80 km/h",
                                         "90 km/h", "100 km/h", "110 km/h", "Unknown"))
ggplot(df_merged, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r 311-row2, out.width='25%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
ggplot(df_merged, aes(x = `Surface condition`)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Distribution of Surface Condition", x = "Surface Condition", y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

ggplot(df_merged, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of Traffic Unit Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = fct_infreq(Manoeuvre))) +
    geom_bar(fill = "coral") +
    labs(title = "Distribution of Manoeuvre", x = "Manoeuvre", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r 311-row3, out.width='50%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
ggplot(df_merged, aes(x = `TU role in first impact`)) +
    geom_bar(fill = "gold") +
    labs(title = "Distribution of TU Role in First Impact", x = "Role", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_tu <- df_merged %>%
    mutate(`TU group` = ifelse(`No. of traffic units involved` >= 6, "6+",
                                as.character(`No. of traffic units involved`)))
df_tu$`TU group` <- factor(df_tu$`TU group`, levels = c("1", "2", "3", "4", "5", "6+"))
ggplot(df_tu, aes(x = `TU group`)) +
  geom_bar(fill = "slateblue") +
  labs(title = "Distribution of No. of Traffic Units Involved", x = "No. of Traffic Units", y = "Count") +
  theme_minimal()
```

The data shows significant class imbalance: Injury crashes are most frequent (114,355), followed by Non-casualty (54,191), with Fatal crashes being the minority (2,416). Most crashes occur in fine weather, daylight, and dry road conditions, suggesting traffic volume has more influence than weather on crash frequency. The 50-60 km/h speed zones account for the most crashes, with Sydney metro being the dominant urbanisation category. Car/car derivatives are the most common vehicle type, "Proceeding in lane" is the most frequent manoeuvre (~100,000), with all other manoeuvres being substantially less frequent. Key traffic units account for the largest share of roles in the first impact, and 2-vehicle crashes are the most common collision type.

### 3.1.2 Missing Values
```{r na check, results='hide'}
colSums(is.na(df_merged))
```

Core modelling variables (Weather, Speed limit, Natural lighting, Degree of crash) have no missing values. Primary hazardous feature (168,803), Primary temporary feature (168,406), and Object hit 2 (165,996) have very high missingness due to their conditional nature and will be dropped during preprocessing.

### 3.1.3 Outliers for Key Variables
```{r summary of TU group, results='hide'}
summary(df_merged$`No. of traffic units involved`)
```

```{r boxplot-tu, fig.width=4, fig.height=3}
ggplot(df_merged, aes(y = `No. of traffic units involved`)) +
  geom_boxplot(fill = "steelblue", outlier.colour = "red", outlier.size = 1) +
  scale_y_log10() +
  labs(title = "Boxplot of No. of Traffic Units Involved (Log Scale)",
       y = "No. of Traffic Units (log scale)") +
  theme_minimal()
```

The number of traffic units has a median of 2 and mean of 2.149, with IQR concentrated between 1 and 2. The maximum value of 26 is an extreme outlier and may be capped during preprocessing.

## 3.2 Bivariate Analysis
### 3.2.1 Features vs Outcomes
```{r 321-row1, out.width='25%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
ggplot(df_merged, aes(x = Weather, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Weather", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = `Speed limit`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Speed Limit", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = `Natural lighting`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Natural Lighting", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = `TU type group`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Type", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r 321-row2, out.width='25%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
ggplot(df_merged, aes(x = fct_infreq(factor(Manoeuvre)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Manoeuvre", x = "Manoeuvre", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

ggplot(df_merged, aes(x = fct_infreq(factor(Urbanisation)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Urbanisation", x = "Urbanisation", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(df_merged, aes(x = fct_infreq(factor(`Surface condition`)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Surface Condition", x = "Surface Condition", y = "Proportion") +
  theme_minimal()

ggplot(df_merged, aes(x = `TU controlled flag`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Controlled Flag", x = "Controlled", y = "Proportion") +
  theme_minimal()
```

Speed limit shows the strongest relationship with crash severity, with higher speed zones (100-110 km/h) having noticeably greater Fatal proportions. Pedestrians and motorcycles show elevated Injury proportions, while uncontrolled traffic units have higher Non-casualty proportions. Country non-urban areas show slightly higher Fatal proportions compared to urban areas, consistent with higher rural speed limits.

### 3.2.2 Features vs Features
```{r 322-row1, out.width='33%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
df_merged %>%
  group_by(Weather, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Natural Lighting", fill = "Count") +
  theme_minimal()

df_merged %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Urbanisation, y = `Speed limit`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Speed Limit vs Urbanisation", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_merged %>%
  group_by(Weather, `Surface condition`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Surface condition` != "Unknown") %>%
  ggplot(aes(x = `Surface condition`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Surface Condition", fill = "Count") +
  theme_minimal()
```
```{r 322-row2, out.width='50%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
df_merged %>%
  group_by(`TU type group`, Manoeuvre) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Manoeuvre, y = `TU type group`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "TU Type vs Manoeuvre", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

df_merged %>%
  group_by(Urbanisation, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(`Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Urbanisation, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Urbanisation vs Natural Lighting", fill = "Count") +
  theme_minimal()
```

Fine weather strongly aligns with dry surfaces while rain almost exclusively pairs with wet conditions, confirming dataset consistency. Speed limits of 50-60 km/h dominate Sydney metro areas while 100-110 km/h zones concentrate in Country non-urban regions. Crashes predominantly occur in daylight across all urbanisation categories, with Sydney metro having the highest overall crash volume.

## 3.3 Additional Visualisations
To further support the exploratory analysis, additional visualizations and summary tables were produced to reveal temporal, spatial, and contextual patterns in crash severity. These displays go beyond simple univariate and bivariate summaries by highlighting high-risk combinations of conditions and identifying patterns that may influence feature selection and model choice.

### 3.3.1 Crash Frequency by Day & Time
```{r eda-helpers, message=FALSE, warning=FALSE}
library(knitr)

# Ensure ordered factors for time-related variables
day_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
df_merged$`Day of week of crash` <- factor(df_merged$`Day of week of crash`, levels = day_levels)

# If Two-hour intervals is character, keep original order as it appears
df_merged$`Two-hour intervals` <- factor(df_merged$`Two-hour intervals`,
                                         levels = unique(df_merged$`Two-hour intervals`))

# Order reporting year if needed
df_merged$`Reporting year` <- factor(df_merged$`Reporting year`,
                                     levels = sort(unique(df_merged$`Reporting year`)))
```
```{r heatmap-day-time, fig.width=8, fig.height=3.5}
df_merged %>%
  group_by(`Day of week of crash`, `Two-hour intervals`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = `Two-hour intervals`, y = `Day of week of crash`, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Crash Frequency by Day of Week and Time Interval",
    x = "Two-hour Interval",
    y = "Day of Week",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Crash frequency peaks during daytime and commuting hours, suggesting traffic exposure drives occurrence more than day of week alone. Time-related variables are likely to be useful predictors.

### 3.3.2 Crash Severity by Year
```{r severity-by-year, fig.width=5, fig.height=3}
df_merged %>%
  group_by(`Reporting year`, `Degree of crash`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = `Reporting year`, y = count, fill = `Degree of crash`)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Crash Severity by Reporting Year",
    x = "Reporting Year",
    y = "Proportion",
    fill = "Crash Severity"
  ) +
  theme_minimal()
```
Severity proportions are stable across years, confirming that class imbalance is a persistent structural feature of the dataset rather than an anomaly, reinforcing the need for imbalance-aware modelling.

### 3.3.3 Fatal Rate by Speed & Urbanisation
```{r fatal-rate-speed-urban, fig.width=6, fig.height=4}
fatal_heatmap <- df_merged %>%
  filter(!is.na(`Speed limit`), !is.na(Urbanisation), !is.na(`Degree of crash`)) %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(
    total = n(),
    fatal_rate = mean(`Degree of crash` == "Fatal"),
    .groups = "drop"
  ) %>%
  filter(total >= 100)  # avoid unstable rates from tiny groups

ggplot(fatal_heatmap, aes(x = Urbanisation, y = `Speed limit`, fill = fatal_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkred",
                      labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Fatal Crash Proportion by Speed Limit and Urbanisation",
    x = "Urbanisation",
    y = "Speed Limit",
    fill = "Fatal Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Fatal crash proportions are highest in high-speed, less-urbanised zones, confirming that speed environment and spatial context jointly influence severity and motivating the use of interaction-capable models.

### 3.3.4 High-Risk Crash Scenarios
```{r high-risk-scenarios-table}
risk_table <- df_merged %>%
  filter(
    !is.na(`Speed limit`),
    !is.na(`Natural lighting`),
    !is.na(Urbanisation),
    !is.na(`Degree of crash`)
  ) %>%
  group_by(`Speed limit`, `Natural lighting`, Urbanisation) %>%
  summarise(
    total_crashes = n(),
    fatal_prop = mean(`Degree of crash` == "Fatal"),
    injury_prop = mean(`Degree of crash` == "Injury"),
    non_casualty_prop = mean(`Degree of crash` == "Non-casualty (towaway)"),
    .groups = "drop"
  ) %>%
  filter(total_crashes >= 200) %>%   # keep only meaningful groups
  arrange(desc(fatal_prop), desc(total_crashes)) %>%   # sort by fatal proportion descending
  slice_head(n = 10) %>%
  mutate(
    fatal_prop = scales::percent(fatal_prop, accuracy = 0.1),
    injury_prop = scales::percent(injury_prop, accuracy = 0.1),
    non_casualty_prop = scales::percent(non_casualty_prop, accuracy = 0.1)
  ) %>%
  rename(
    `Speed Limit` = `Speed limit`,
    `Natural Lighting` = `Natural lighting`,
    `Total Crashes` = total_crashes,
    `Fatal Proportion` = fatal_prop,
    `Injury Proportion` = injury_prop,
    `Non-casualty Proportion` = non_casualty_prop
  )

knitr::kable(
  risk_table,
  caption = "Table: High-risk crash scenarios ranked by fatal crash proportion"
)
```
High fatal proportions cluster around combinations of high speed limits, darkness, and rural contexts, confirming that crash severity is shaped by interacting conditions rather than any single variable.

### 3.3.5 Top 10 LGAs & Fatal Rate by Traffic Unit Type
```{r lga-tu-pair, out.width='50%', fig.show='hold', fig.height=4, message=FALSE, warning=FALSE}
top_lga <- df_merged %>%
  filter(!is.na(LGA)) %>%
  count(LGA, sort = TRUE) %>%
  slice_head(n = 10)

ggplot(top_lga, aes(x = reorder(LGA, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 LGAs by Crash Count", x = "LGA", y = "Number of Crashes") +
  theme_minimal()

tu_fatal <- df_merged %>%
  filter(!is.na(`TU type group`), !is.na(`Degree of crash`)) %>%
  group_by(`TU type group`) %>%
  summarise(total = n(), fatal_rate = mean(`Degree of crash` == "Fatal"), .groups = "drop") %>%
  filter(total >= 100) %>%
  arrange(desc(fatal_rate))

ggplot(tu_fatal, aes(x = reorder(`TU type group`, fatal_rate), y = fatal_rate)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Fatal Crash Proportion by Traffic Unit Type",
       x = "Traffic Unit Type", y = "Fatal Proportion") +
  theme_minimal()
```
Crash counts are spatially concentrated in a few LGAs, and fatal proportions vary substantially across traffic unit types — both reinforcing the predictive value of location and vehicle-level features.


# 4 Project Planning
## 4.1 Preprocessing Plan

To transform the raw NSW Road Crash Data into a clean, model-ready dataset, we propose the following preprocessing pipeline. All data-dependent steps (imputation, encoding, scaling, resampling) will be performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

| Step | Action | Key Details |
|------|--------|-------------|
| 1. Data Integration | Aggregate TU → crash level; left-join on `Crash ID` | Binary TU flags (`has_heavy_vehicle`, `has_motorcycle`, etc.); TU type count |
| 2. Feature Removal | Drop leakage, identifiers, redundant & sparse columns | Remove `No. killed`, injury counts (leakage); `Street`, `Route no.` (redundant); columns >85% NA |
| 3. Missing Values | Targeted imputation per variable | High-NA cols → binary flags; `Other TU type` → `"Single vehicle"`; categoricals → `"Unknown"`; numeric → median |
| 4. Feature Engineering | Derive binary & grouped features | `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`; speed bands (≤50, 60–80, 90–100, 110+) |
| 5. Outlier Detection | Remove geospatial & distance outliers | Drop outside NSW bounds; winsorise `Distance` at 99th percentile |
| 6. Encoding | Model-dependent categorical encoding | Trees: R `factor`; kNN/SVM/LR: one-hot/ordinal; pool rare `LGA` (<1%) into `"Other"` |
| 7. Scaling | Standardise numerics for distance-based models | Z-score within fold for kNN, SVM; trees scale-invariant — no scaling needed |
| 8. Class Imbalance | Resample within training fold only | SMOTE for Fatal minority class; compare with down-sampling during tuning (Chawla et al., 2002) |
| 9. Train/Test Split | 80/20 stratified split; nested 5-fold CV | All steps in `tidymodels::recipe`; transformations fitted on training data only per fold |

## 4.2 Model Plan: Proposed Classification Models

Based on the findings from EDA, we propose six classification models to predict **Degree of Crash** (a three-class outcome: Fatal, Injury, and Non-casualty). The EDA revealed that the target variable is extremely imbalanced—Fatal crashes account for a small proportion of the 92,189 observations—and that all predictor variables are categorical (e.g., `Weather`, `Natural lighting`, `Surface condition`, `Speed limit`, `Urbanisation`, `TU type group`). These characteristics directly inform the choice of models below. All models will be implemented in R using the `caret` package for a unified training and cross-validation interface.

| # | Model | Justification | Implementation |
|---|-------|---------------|----------------|
| 1 | Multinomial Logistic Regression (Baseline) | Interpretable baseline; log-odds per predictor meaningful for stakeholders. All models benchmarked against this. | `multinom()` from `nnet`; dummy-encoded predictors; `class_weight` for imbalance. |
| 2 | Decision Tree (CART) | Fully interpretable; captures interactions (e.g., high speed + rural → fatal) in a single branch. | `rpart`; depth tuned via CV; visualised with `rpart.plot`. |
| 3 | Random Forest | Bootstrap ensemble reduces variance; variable importance scores rank predictors. | `randomForest`; `classwt` for imbalance; feasible at >90k obs. |
| 4 | Gradient Boosting (XGBoost) | High-performance; sequential correction of misclassifications effective for rare Fatal class. | `xgboost`; `nrounds`, `max_depth`, `eta` tuned via `caret`; `scale_pos_weight` for imbalance. |
| 5 | K-Nearest Neighbours (KNN) | Non-parametric; tests local consistency — do similar-condition crashes share severity? | `class` + Gower distance for categoricals; optimal $k$ via CV. |
| 6 | Naive Bayes | Natural for all-categorical data; probabilistic benchmark for individual feature signal. | `e1071`; fast under conditional independence; outputs class probabilities. |

## 4.3 Evaluation Plan: Performance Metrics

Standard accuracy is an inadequate evaluation metric for this problem due to the class imbalance in Degree of Crash. At the vehicle level, Fatal observations account for only 1.4% of the 170,962 records, with Injury at 66.9% and Non-casualty at 31.7%. A classifier that predicts Injury for every observation would achieve high accuracy simply because Injury is the most frequent class, while completely failing to identify any Fatal crash. The metrics below are selected to reflect both overall classifier performance and the disproportionate importance of correctly identifying fatal outcomes.

| Metric | Role | Justification |
|--------|------|---------------|
| Macro-Averaged F1 Score | Primary | Equal-weight F1 across all 3 classes; ensures Fatal class not ignored despite low frequency. Primary model selection metric. |
| Fatal Class Recall | Primary Constraint | % of actual Fatal crashes correctly flagged; hard constraint — low recall deprioritises the model regardless of overall F1. |
| Weighted F1 Score | Secondary | F1 weighted by class support; reflects real-world class distribution performance alongside macro F1. |
| Cohen's Kappa (κ) | Secondary | Chance-adjusted accuracy; near-zero = no better than random. Via `caret`'s `confusionMatrix()`. |
| Confusion Matrix | Secondary | 3×3 heatmap showing misclassification direction — e.g., Fatal predicted as Injury (most dangerous error). |
| Cross-Validation Strategy | Validation | Grouped 5-fold CV by `Crash ID`; temporal split: 2020–2023 train, 2024 test; test set evaluated once only. |

## 4.4 Timeline and Project Milestones

The project timeline has been structured in alignment with the assignment deadlines to ensure clear and systematic progression of tasks from understanding the dataset up to final model development and reporting.

| Period | Milestone | Key Activities |
|--------|-----------|----------------|
| 1–7 April | Problem Framing & Data Acquisition | Sourced & selected dataset; finalised research question. |
| 7–10 April | Data Integration & Initial Preparation | Merged datasets on Crash ID (170,962 records); examined missingness and variable types. |
| 10–14 April | Exploratory Data Analysis (EDA) | Univariate & bivariate analyses done; insights used for feature selection. |
| 14–17 April | Feature Planning & Model Design | Features identified from EDA; encoding plan and model list finalised. |
| 17–19 April | Plan & EDA Report Finalisation | Report compiled and submitted. |
| 6–10 May | Data Preprocessing & Baseline Modelling | Preprocessing pipeline implemented; Logistic Regression baseline trained and evaluated. |
| 11–15 May | Advanced Model Development & Tuning | Remaining 5 models implemented and hyperparameter-tuned via CV. |
| 16–20 May | Model Evaluation & Report Writing | Models evaluated; final model selected; results consolidated into report. |
| 21–24 May | Report Refinement | Final edits for clarity and submission requirements. |
| 25–31 May | Final Checks & Submission | Contributions reviewed; final check completed before submission. |

**Backup Plan:** The following contingency measures have been put in place. Two members are on standby at any time to cover tasks if a member is unable to meet a deadline; all work is stored in a shared GitHub repository. The team holds two meetings per week (Thursday in-person, Monday via Zoom) with 2 check-ins per week to monitor progress and flag delays early. Each member contributes approximately 20% across both phases.

# References

Chawla, N. V., Bowyer, K. W., Hall, L. O., & Kegelmeyer, W. P. (2002). SMOTE: Synthetic minority over-sampling technique. *Journal of Artificial Intelligence Research*, *16*, 321–357. https://doi.org/10.1613/jair.953

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning: With applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Little, R. J. A. (1988). A test of missing completely at random for multivariate data with missing values. *Journal of the American Statistical Association*, *83*(404), 1198–1202. https://doi.org/10.1080/01621459.1988.10478722

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. Australian Government Open Data Platform. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety. https://www.transport.nsw.gov.au/system/files/media/documents/2025/crs_road_safety_progress_report_2024.pdf

# Acknowledgement of AI

Group 07 (Workshop 12) declares that all ideas, analysis, and written content 
presented in this report are our own. We did not use generative AI to write our 
code. We used ChatGPT (GPT-5.3, OpenAI, https://chatgpt.com) and Claude Sonnet 4.5 
(Anthropic, https://claude.ai) for code refinement, grammar checking, and 
reviewing our report against the marking rubric.
