---
title: "NSW Road Crash Severity Prediction: Project Plan and EDA"
author: "Group 7, Workshop 12"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(readxl)
library(forcats)
library(scales)
library(knitr)
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.height = 3, fig.width = 6)
df_merged <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
df_merged$`Speed limit` <- factor(df_merged$`Speed limit`,
  levels = c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h","60 km/h",
             "70 km/h","80 km/h","90 km/h","100 km/h","110 km/h","Unknown"))
day_levels <- c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")
df_merged$`Day of week of crash` <- factor(df_merged$`Day of week of crash`, levels = day_levels)
df_merged$`Two-hour intervals` <- factor(df_merged$`Two-hour intervals`,
                                         levels = unique(df_merged$`Two-hour intervals`))
df_merged$`Reporting year` <- factor(df_merged$`Reporting year`,
                                     levels = sort(unique(df_merged$`Reporting year`)))
```

# 1 Introduction & Research Question

Given environmental conditions (weather, natural lighting, road surface, speed limit) and traffic unit characteristics (number of traffic units involved, type of traffic units, road classification), can we predict the severity of a road crash in NSW as a multi-class classification problem: **Fatal**, **Injury**, or **Non-casualty**?

This question has direct real-world relevance — NSW recorded 340 road fatalities in 2024, with over two-thirds in regional areas (Transport for NSW, 2025). Accurate severity prediction supports targeted resource allocation for traffic authorities, emergency response teams, and policymakers.

# 2 Data

The dataset is sourced from Transport for NSW via the Australian Government open data platform. Two tables were merged on `Crash ID`:

- **CRASH** (92,189 records): environmental conditions, road characteristics, crash outcomes
- **TRAFFIC UNIT** (170,962 records): vehicle type, manoeuvre, role in first impact

The merged dataset contains **170,962 records at the vehicle level**, providing rich environmental and traffic-unit features for modelling. It satisfies the assignment requirements: large (>10,000 observations), integrated (multiple sources), messy (real-world missing values and duplication), and supports a multi-class classification target.

**Missing values:** Core modelling variables (Weather, Speed limit, Natural lighting, Degree of crash) have no missing values. Primary hazardous feature (98.5% missing), Primary temporary feature (98.4% missing), and Object hit 2 (97.1% missing) will be dropped during preprocessing due to their conditional and sparse nature.

# 3 Exploratory Data Analysis

## 3.1 Univariate Analysis

```{r fig.height=2.5, fig.width=5}
ggplot(df_merged, aes(x = fct_infreq(`Degree of crash`))) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
  theme_minimal()
```

```{r fig.height=2.2, fig.width=10}
p1 <- ggplot(df_merged, aes(x = fct_infreq(Weather))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Weather", x = NULL, y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)

p2 <- ggplot(df_merged, aes(x = fct_infreq(`Natural lighting`))) +
  geom_bar(fill = "orange") +
  labs(title = "Natural Lighting", x = NULL, y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(df_merged, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Speed Limit", x = NULL, y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(df_merged, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Urbanisation", x = NULL, y = "Count") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 1)
```

The data shows significant class imbalance: Injury crashes are most frequent (114,355), followed by Non-casualty (54,191), with Fatal crashes being the minority (2,416). Most crashes occur in fine weather, daylight, and dry road conditions, suggesting traffic volume has more influence than weather on crash frequency. The 50–60 km/h speed zones account for the most crashes, with Sydney metro being the dominant urbanisation category. Car/car derivatives are the most common vehicle type, "Proceeding in lane" is the most frequent manoeuvre (~100,000). Key traffic units account for the largest share of roles in the first impact, and 2-vehicle crashes are the most common collision type.

## 3.2 Bivariate Analysis: Features vs Outcome

```{r fig.height=2.5, fig.width=10}
b1 <- ggplot(df_merged, aes(x = `Speed limit`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Severity by Speed Limit", y = "Proportion", x = NULL) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1),
                          legend.position = "none")

b2 <- ggplot(df_merged, aes(x = `Natural lighting`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Severity by Lighting", y = "Proportion", x = NULL) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1),
                          legend.position = "none")

b3 <- ggplot(df_merged, aes(x = fct_infreq(factor(Urbanisation)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Severity by Urbanisation", y = "Proportion", x = NULL) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1),
                          legend.key.size = unit(0.3, "cm"))

gridExtra::grid.arrange(b1, b2, b3, nrow = 1)
```

Speed limit shows the strongest relationship with crash severity — higher speed zones (100–110 km/h) have noticeably greater Fatal proportions. Pedestrians and motorcycles show elevated Injury proportions, while uncontrolled traffic units have higher Non-casualty proportions. Country non-urban areas show slightly higher Fatal proportions, consistent with higher rural speed limits.

## 3.3 Feature Interactions & High-Risk Scenarios

```{r fig.height=2.5, fig.width=5}
fatal_heatmap <- df_merged %>%
  filter(!is.na(`Speed limit`), !is.na(Urbanisation), !is.na(`Degree of crash`)) %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(total = n(), fatal_rate = mean(`Degree of crash` == "Fatal"), .groups = "drop") %>%
  filter(total >= 100)

ggplot(fatal_heatmap, aes(x = Urbanisation, y = `Speed limit`, fill = fatal_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkred",
                      labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Fatal Proportion by Speed Limit and Urbanisation",
       x = NULL, y = NULL, fill = "Fatal %") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
risk_table <- df_merged %>%
  filter(!is.na(`Speed limit`), !is.na(`Natural lighting`),
         !is.na(Urbanisation), !is.na(`Degree of crash`)) %>%
  group_by(`Speed limit`, `Natural lighting`, Urbanisation) %>%
  summarise(total_crashes = n(),
            fatal_prop = mean(`Degree of crash` == "Fatal"), .groups = "drop") %>%
  filter(total_crashes >= 200) %>%
  arrange(desc(fatal_prop)) %>%
  slice_head(n = 6) %>%
  mutate(fatal_prop = scales::percent(fatal_prop, accuracy = 0.1)) %>%
  rename(`Speed Limit` = `Speed limit`, `Natural Lighting` = `Natural lighting`,
         `Total Crashes` = total_crashes, `Fatal %` = fatal_prop)

knitr::kable(risk_table, caption = "Top high-risk crash scenarios by fatal proportion")
```

Fatal crash proportions are highest in high-speed, less-urbanised regions, particularly at night. These interaction patterns motivate the use of flexible, interaction-capable models. Fine weather strongly aligns with dry surfaces while rain almost exclusively pairs with wet conditions, confirming dataset consistency. Speed limits of 50–60 km/h dominate Sydney metro areas while 100–110 km/h zones concentrate in Country non-urban regions.

# 4 Project Plan

## 4.1 Preprocessing Pipeline

All data-dependent steps will be performed **within cross-validation folds** to prevent data leakage.

| Step | Action |
|------|--------|
| **Integration** | Aggregate TRAFFIC UNIT to crash level (binary flags for vehicle types); left-join to CRASH on `Crash ID` |
| **Feature removal** | Drop identifiers, target-leakage variables (No. killed, No. injured), near-zero-variance columns (>95% missing) |
| **Missing values** | `Primary permanent feature` → binary indicator; `Other TU type` NA → "Single vehicle"; remaining categoricals → "Unknown" level |
| **Feature engineering** | `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, speed bands (≤50, 60–80, 90–100, 110+) |
| **Encoding** | One-hot for linear/distance models; native factors for tree-based models |
| **Scaling** | Z-score standardisation for kNN and SVM only |
| **Class imbalance** | SMOTE via `themis::step_smote` within training fold; compare with down-sampling |

## 4.2 Models

| Model | Justification |
|-------|--------------|
| **Multinomial Logistic Regression** | Interpretable baseline; log-odds coefficients meaningful for stakeholders |
| **Decision Tree (CART)** | Captures non-linear interactions; fully interpretable via tree visualisation |
| **Random Forest** | Reduces variance via ensembling; provides variable importance scores |
| **XGBoost** | High-performance boosting; effective at learning from the rare Fatal class |
| **K-Nearest Neighbours** | Tests local consistency of crash severity in feature space |
| **Naive Bayes** | Probabilistic benchmark suited to all-categorical features |

## 4.3 Evaluation

Standard accuracy is inadequate due to severe class imbalance (Fatal: 1.5%, Injury: 66.4%, Non-casualty: 32.1%). Primary metric is **Macro-Averaged F1**, ensuring equal weight to the rare Fatal class. **Fatal Class Recall** is a primary constraint — models with high Macro F1 but low Fatal recall will be deprioritised. Secondary metrics include Weighted F1 and Cohen's Kappa. Confusion matrices will be visualised to monitor dangerous misclassifications (Fatal predicted as Injury).

**Validation:** Stratified 5-fold CV on training data. Temporal holdout: 2020–2023 for training, 2024 as held-out test set. Final test set evaluated once only after model selection.

## 4.4 Timeline

| Period | Milestone |
|--------|-----------|
| 1–10 Apr | Problem framing, data acquisition, integration |
| 10–17 Apr | EDA and feature planning |
| 17–19 Apr | Plan & EDA report submission |
| 6–15 May | Preprocessing, baseline and advanced modelling |
| 16–20 May | Model evaluation and report writing |
| 21–31 May | Refinement, final checks, submission |

# References

Chawla et al. (2002). SMOTE. *JAIR*, *16*, 321–357.

James et al. (2021). *An introduction to statistical learning* (2nd ed.). Springer.

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety.

# Acknowledgement of AI

Group 07 (Workshop 12) declares that all ideas, analysis, and written content are our own. We used ChatGPT (GPT-5.3) and Claude Sonnet 4.5 for code refinement, grammar checking, and reviewing our report against the marking rubric.
