---
title: "STAT5003 Assignment Title"
author: "W12G07"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---
# Introduction

```{r library, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(readxl)  
library(forcats)
```

```{r read crash dataset}
df_crash <- read_xlsx("nsw_road_crash_data_2020-2024_crash.xlsx")
str(df_crash)
head(df_crash)
```

```{r read traffic unit}
df_unit <- read_xlsx("nsw_road_crash_data_2020-2024_traffic_unit.xlsx")
str(df_unit)
head(df_unit)
```


```{r merge & save}
df_merged <- left_join(df_crash, df_unit, by = "Crash ID")
str(df_merged)
head(df_merged)
write.csv(df_merged, "nsw_road_crash_merged_2020_2024.csv", row.names = FALSE)
```

# 1 Framing & Dataset Selection

## 1.1 Classification Task

Based on environmental conditions and traffic unit characteristics, predict the 
severity of a road crash as a multi-class classification problem: Fatal, 
Injury, or Non-casualty.

                                                      
## 1.2 Research Question

Given environmental conditions (weather, natural lighting, road surface, 
speed limit) and traffic unit characteristics (number of traffic units involved, 
type of traffic units, road classification), can we predict the severity 
of a road crash in NSW?                                                        
  
## 1.3 Motivation                                                

This research question has direct real-world relevance and affects the following 
groups:
  
NSW Traffic Management Authorities: By identifying key combinations of 
environmental and traffic characteristics that contribute to severe crashes, 
authorities can make targeted decisions such as adjusting speed limits, 
improving road design, and increasing enforcement during high-risk conditions.   

Emergency Response Teams: In the early stages of a crash, predicting severity 
based on environmental conditions can help allocate ambulance and emergency 
resources more quickly and effectively.

Policymakers and Insurance Institutions: Understanding which traffic unit types 
(e.g. heavy vehicles) pose higher risks under specific environmental conditions 
can inform more precise road safety policies and insurance pricing models.

The dataset contains 92,189 crash records across NSW from 2020 to 2024, covering 
variables such as weather, natural lighting, road surface condition, speed 
limit, number of traffic units involved, and traffic unit types: fully 
supporting the environmental and traffic unit characteristics required by our 
research question.

A predictive model for crash severity enables data-driven prioritization, this 
can help authorities allocate resources design intervention strategies, and form 
policy based on evidence rather than intuition. 

# 2 Consider & Gather Data 
```{r}
cat("Merged:", nrow(df_merged), "rows,", ncol(df_merged), "cols\n")
colSums(is.na(df_merged))
```
```{r na visualisation}
na_counts <- colSums(is.na(df_merged))
na_df <- data.frame(
  variable = names(na_counts),
  missing = na_counts
) |> subset(missing > 0)

ggplot(na_df, aes(x = reorder(variable, -missing), y = missing)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(title = "Figure 1: Missing Values by Column", x = "Column", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## 2.1 Data Source 

The data used in this study is sourced from the Australian Government open data
platform, **NSW Crash Data Dataset:** 
https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data 

This dataset is published by Transport for NSW and provides detailed information 
on road crashes in NSW, including environmental conditions, road characteristics
, vehicle types, and injury outcomes 

For this study, we utilized the Crash Dataset (NSW Road Crash Data - 2020-2024 
- CRASH.xlsx) and Traffic Unit Dataset (NSW Road Crash Data - 2020-2024 - 
TRAFFIC UNIT.xlsx ) for 2020-2024.The crash dataset contains environmental and 
road- related variables (e.g. weather, road surface, and speed limit ) 

- The crash dataset contains: 
  - Crash Details (No. Killed, No. Seriously Injured, Severity of Crash) 
  - time-related features (Day, Month, etc) 
  - location attributes (Direction, latitude, longitude, Distance, etc) 
  - environmental conditions (Weather, Natural Lighting, Spped Limit, etc.) 
  
- The traffic unit dataset contains vehicle-level variables: 
  - TU type group (states vehicle type) 
  - maneuver (states vehicle movement at time of crash) 
  - TU role in first impact (states role of vehicle e.g. key traffic unit) 
  - TU controlled flag (Whether classified as controlled or uncontrolled)
  
## 2.2 Dataset Overview

The crash dataset contains 92,189 crash records, and the traffic dataset 
contains 170,962 records across NSW from 2020 to 2024. The dataset formed by 
merging both of these on Crash ID contains 170,962 records. The merged dataset 
is recorded at a vehicle level. This allows vehicle-specific characteristics to 
be incorporated into the crash severity increasing the richness of our dataset.

## 2.3 Dataset Justification

The dataset meets several key requirements outlined in the assignment.

**Large Dataset(>10,000 observations)**

The merged dataset has 170,962 records, providing sufficient scale for robust 
statistical analysis and model training. The large size enables reliable 
estimation of relationships between our target variable (crash severity) and its 
predictors while also supporting validation techniques such as test-train split.

**Integrated Dataset (Multiple sources combined)**

The dataset is formed by merging crash-level and traffic unit datasets using a 
common Crash ID. This integration combines environmental, spatial, and vehicle-
level information. The process improves the richness of features available for 
analysis and enables more comprehensive modelling of crash severity.

**Messy Dataset (Real-world data challenge)**

As a real-world administrative dataset, it contains missing values and potential 
inconsistencies across variables such as weather, lighting, and vehicle 
characteristics. Additionally, the process of merging the crash dataset (92,189 
records) with traffic unit dataset (170,962 records) exposes existing missing
values, as not all variables are available at both the crash and vehicle levels. 
When merged using Crash ID, crash-level attributes are duplicated across 
multiple vehicle records, while some vehicle-specific fields may not have 
corresponding values for all observations.

This results in a dataset with increased sparsity and inconsistencies across 
columns, requiring preprocessing steps such as handling missing data, managing 
duplicated information, and cleaning categorical variables. As shown in the 
above figure, columns such as Primary hazardous feature, Primary temporary 
feature, and object hit 2 have the highest missingness(>95%) while Route no. and 
Other TU type have moderate missingness

**Multi-class Classification Problem**

The target variable, crash severity, is a multi-class outcome derived from 
Degree of crash column in the crash dataset, consisting of  categories such as 
Fatal, Injury, and Non-casualty. These categories are supported by injury-related 
variables in the crash dataset (e.g., number killed and 
number seriously injured). This introduces challenges such as class imbalance, 
which must be considered when evaluating model performance.

**2.4 Summary**

Overall, the dataset is well-suited for this study as it is large, integrated, 
messy, and supports a multi-classification problem. Its combination of 
environmental, temporal, spatial, and vehicle-level features provides a strong 
foundation for modelling and predicting crash severity.
```{r} 
# R code...
```

# 3 Exploratory Data Analysis (EDA)
## 3.1 Univariate Analysis
### 3.1.1 Distribution
```{r barpolot of degree of crash}
ggplot(df_merged, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")
```
```{r barplot of weather}
ggplot(df_merged, aes(x = fct_infreq(Weather))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of Weather Conditions", x = "Weather", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of natrual lighting}
ggplot(df_merged, aes(x = fct_infreq(`Natural lighting`))) +
  geom_bar(fill = "orange") +
  labs(title = "Distribution of Natrual lighting", x = "Lighting", y = "Count") + 
  theme_minimal()
```
```{r bar plot of speed limit}
df_merged$`Speed limit` <- factor(df_merged$`Speed limit`,
                              levels = c("10 km/h", "20 km/h", "30 km/h", "40 km/h",
                                         "50 km/h", "60 km/h", "70 km/h", "80 km/h",
                                         "90 km/h", "100 km/h", "110 km/h", "Unknown"))

  ggplot(df_merged, aes(x = `Speed limit`)) +
    geom_bar(fill = "tomato") +
    labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of Surface condition}
ggplot(df_merged, aes(x = `Surface condition`)) + 
  geom_bar(fill = "darkgreen") +
  labs(title = "Distribution of Surface condition", x = "Surface condition", y = "Count") +
  theme_minimal()
```
```{r barplot of Urbanisation}
ggplot(df_merged, aes(x = fct_infreq(Urbanisation))) + 
  geom_bar(fill = "purple") + 
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r bar plot of TU type group}
ggplot(df_merged, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of Traffic Unit Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of Manoeuvre}
ggplot(df_merged, aes(x = fct_infreq(Manoeuvre))) +
    geom_bar(fill = "coral") +
    labs(title = "Distribution of Manoeuvre", x = "Manoeuvre", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of TU role in first impact}
ggplot(df_merged, aes(x = `TU role in first impact`)) +
    geom_bar(fill = "gold") +
    labs(title = "Distribution of TU Role in First Impact", x = "Role", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r barplot of TU group}
df_tu <- df_merged %>%
    mutate(`TU group` = ifelse(`No. of traffic units involved` >= 6, "6+",
                                as.character(`No. of traffic units involved`)))                                                                            
df_tu$`TU group` <- factor(df_tu$`TU group`, levels = c("1", "2", "3", "4", "5", "6+"))                                                                  
ggplot(df_tu, aes(x = `TU group`)) +                                                                                                                     
geom_bar(fill = "slateblue") +                          
labs(title = "Distribution of No. of Traffic Units Involved", x = "No. of Traffic Units", y = "Count") + theme_minimal() 
```
The distribution of crash severity reveals significant class imbalance, we can 
see Injury crashes being the most common (114,355) followed by Non-casualty 
(54,191), and Fatal crashes being in minority (2,416). The majority of crashes 
are occurring during fine weather (>150,000) in daylight conditions, suggestive
of higher traffic volume rather than driving conditions like visibility drive 
overall crash frequency. Road surfaces were also predominantly dry, consistent
with fine weather dominance. Speed limits of 50 km/h and 60 km/h accounted for 
highest number of crashes, reflecting urban driving environments, this is also 
supported by Sydney metro area being the most common urbanization category. 
Car/car derivative is by far the most common traffic unit type (~115,000), 
followed by light trucks and motorcycles. The most frequent manoeuvre was 
"Proceeding in lane" (~1000,000), with all other manoeuvres being substantially
less frequent, indicating highly skewed data. The majority of traffic units were
classified as Key traffic units in the first impact. The number of traffic units 
involved is right-skewed, with 2-vehicle crashes being the most common, followed
by single and 3-vehicle crashes. Crashes involving 4 or more vehicles are rare.
### 3.1.2 missing values
```{r na check}
colSums(is.na(df_merged))
```
The missingness analysis confirms the core variables used for modelling 
(Weather, Speed limit, Natural lighting, Degree of crash) have no missing values.
High missingness in Primary hazardous feature (168,803), Primary temporary 
feature (168,406), and object hit 2 (165,996) reflects their conditional nature,
these fields are only populated when applicable. We will drop these columns 
during pre-processing. Route no (62,005) and Other TU type (29,195) have 
moderate missingness and will require further consideration.

### 3.1.3 outliers for key variables.
```{r summary of TU group}
summary(df_merged$`No. of traffic units involved`)
```

```{r}
ggplot(df_merged, aes(y = `No. of traffic units involved`)) +
  geom_boxplot(fill = "steelblue", outlier.colour = "red", outlier.size = 1) +
  scale_y_log10() +
  labs(title = "Boxplot of No. of Traffic Units Involved (Log Scale)",
       y = "No. of Traffic Units (log scale)") +
  theme_minimal()
```
The number of traffic units involved has a median 2 and mean 2.149, with IQR 
concentrated between 1 and 2. The boxplot reveals multiple outliers above 3, 
with a maximum of 26 which represents a large multi-vehicle crash that is 
statistically rare. These extreame values may be capped or treated separately 
during preprocessing.

## 3.2 Bivariate Analysis
### 3.2.1 Features vs Outcomes
```{r weather vs degree of crash}
ggplot(df_merged, aes(x = Weather, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Weather", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r speed limit vs degree of crash}
ggplot(df_merged, aes(x = `Speed limit`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Speed Limit", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r natural lighting vs degree of crash}
ggplot(df_merged, aes(x = `Natural lighting`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Natural Lighting", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r TU type group vs degree of crash}
ggplot(df_merged, aes(x = `TU type group`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Type", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r manoeuvre vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(Manoeuvre)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Manoeuvre", x = "Manoeuvre", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```
```{r Urbanisation vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(Urbanisation)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Urbanisation", x = "Urbanisation", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r Surface condition vs degree of crash}
ggplot(df_merged, aes(x = fct_infreq(factor(`Surface condition`)), fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by Surface condition", x = "Surface condition", y = "Proportion") +
  theme_minimal()
```
```{r TU controlled flag vs degree of crash}
ggplot(df_merged, aes(x = `TU controlled flag`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  labs(title = "Crash Severity by TU Controlled Flag", x = "Controlled", y = "Proportion") +
  theme_minimal()
```
Speed limits show the strongest relationship with crash severity, higher speed 
zones (100-110 km/h) have noticeable greater proportions of Fatal crashes, while
lower speed zones (10-30 km/h) are dominated by Injury and Non-casualty outcomes. 
Natural lighting shows slightly higher association of Darkness than Daylight for 
Fatal outcomes. Weather and surface show uniform severity distributions across 
categories, suggesting limited discriminative power. Pedestrians and motorcycles
show a high proportion of Injury crashes relative to other vehicle types, while 
non-motorized vehicles show a notably high Non-casualty proportion (75%). TU 
controlled flag shows a clear difference, uncontrolled units (N) have a much
higher Non-casualty proportion (~55%) compared to controlled units (Y), which 
are more associated with Injury crashes. Country non-urban areas show a slightly 
elevated Fatal proportion compared to Sydney metro, consistent with higher speed 
limits in rural areas.  
### 3.2.2 Features vs Features
```{r weather vs natural lighting}
df_merged %>%
  group_by(Weather, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Natural Lighting", fill = "Count") +
  theme_minimal()
```

```{r speed limit vs urbanisation}
df_merged %>%
  group_by(`Speed limit`, Urbanisation) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Urbanisation, y = `Speed limit`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Speed Limit vs Urbanisation", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r Weather vs Surface condition}
df_merged %>%
  group_by(Weather, `Surface condition`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(Weather != "Unknown", `Surface condition` != "Unknown") %>%
  ggplot(aes(x = `Surface condition`, y = Weather, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Weather vs Surface Condition", fill = "Count") +
  theme_minimal()
```

```{r TU type group vs Manoeuvre}
df_merged %>%
  group_by(`TU type group`, Manoeuvre) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = Manoeuvre, y = `TU type group`, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "TU Type vs Manoeuvre", fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```

```{r Urbanisation vs Natural lighting}
df_merged %>%
  group_by(Urbanisation, `Natural lighting`) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(`Natural lighting` != "Unknown") %>%
  ggplot(aes(x = `Natural lighting`, y = Urbanisation, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "purple") +
  labs(title = "Urbanisation vs Natural Lighting", fill = "Count") +
  theme_minimal()
```
Fine weather is strongly concentrated in Daylight and Darkness conditions, while 
Rain occurs across both Daylight and Darkness, confirming expected weather-
lighting interactions. Speed limit is strongly associated with urbanization, 
50-60 km/h zones dominate Sydney metro areas, while 100-110 km/h zones are 
concentrated in Country non-urban regions. Fine weather also shows overwhelming 
paring with dry surface conditions, while Raining aligns almost exclusively with
wet surfaces, confirming internal consistency in our dataset. Car/car 
derivatives proceeding in lane dominates the TU Type vs Manoeuvre heatmap, 
reflecting their overall frequency in the dataset. Sydney metro area has the 
highest concentration of crashes in both Daylight and Darkness conditions, 
reflecting its overall higher crash volume. All urban categories show crashes 
predominantly occurring in Daylight, with Darkness being the second most common 
lighting condition across all areas. 

# 4.Project Planning
## 4.1 Preprocessing Plan for handling your dataset’s specific requirements to make it clean and ready for modelling.

To transform the raw NSW Road Crash Data into a clean, model-ready dataset, we propose the following preprocessing pipeline. All data-dependent steps (imputation, encoding, scaling, resampling) will be performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

### Step 1: Data Integration

The dataset consists of two linked tables — **CRASH** (92,189 rows × 50 columns) and **TRAFFIC UNIT** (170,962 rows × 10 columns) — connected via `Crash ID`. Since this is a one-to-many relationship, we will **aggregate** the TRAFFIC UNIT table to crash level before joining:

-   Binary flags for vehicle type presence: `has_heavy_vehicle`, `has_motorcycle`, `has_pedestrian`, `has_bicycle`
-   Count of distinct TU types per crash
-   Binary flag for whether any object was hit

The aggregated table will be **left-joined** to CRASH on `Crash ID`, producing a single analysis-ready table.

### Step 2: Feature Selection & Removal

| Category | Columns | Reason |
|----|----|----|
| Identifiers | `Crash ID` | Not predictive |
| Target leakage | `Degree of crash - detailed`, `No. killed`, `No. seriously injured`, `No. moderately injured`, `No. minor-other injured` | Direct consequences of severity — would leak the target |
| Redundant location | `Street of crash`, `Identifying feature`, `Town`, `Route no.`, `Street type`, `Direction` | High cardinality; location already captured by `LGA`, `Lat/Lon`, `Urbanisation` |
| Near-zero variance | `Primary temporary feature` (98.5% NA), `Primary hazardous feature` (98.0% NA), `DCA supplement` (86.8% NA) | Too sparse to be informative |

Retained features cover **environmental conditions** (weather, lighting, road surface, speed limit), **road characteristics** (alignment, type of location), **temporal patterns** (month, day of week, time interval), **spatial context** (urbanisation, LGA), and **traffic unit characteristics** (aggregated from TU table).

### Step 3: Missing Value Handling

| Variable | Missing % | Strategy |
|----|----|----|
| `Primary permanent feature` | 73.0% | Convert to binary indicator `has_permanent_feature`; drop original |
| `Other TU type` | 30.4% | Recode `NA` → `"Single vehicle"` (missing implies single-vehicle crash) |
| Remaining categorical | \<1% | Treat `NA` as an explicit `"Unknown"` level |
| Remaining numeric | \<1% | Median imputation (within training fold) |

For categorical variables, treating missing as its own category is justified when the missingness pattern itself is informative (Little, 1988). We will verify this using `naniar::mcar_test()` and visualise with `visdat::vis_miss()`.

### Step 4: Feature Engineering

Derived features to enhance predictive power:

-   **Temporal**: `is_weekend` (Sat/Sun), `is_rush_hour` (7–9am / 4–6pm)
-   **Environmental**: `is_night` (Darkness/Twilight), `is_wet` (non-Dry surface)
-   **Speed bands**: Group `Speed limit` into `≤50`, `60–80`, `90–100`, `110+`, `Other`
-   **Multi-vehicle**: Binary flag from `No. of traffic units involved > 1`

### Step 5: Outlier Detection

-   **Latitude/Longitude**: Remove records outside NSW bounds (lat: −37.5 to −28.0, lon: 140.0 to 154.0)
-   **Distance**: Inspect for extreme values; winsorise at the 99th percentile if needed

### Step 6: Encoding Categorical Variables

Encoding strategy is model-dependent:

-   **Tree-based models** (Random Forest, XGBoost): Use R `factor` type directly — no encoding needed
-   **Distance/linear models** (kNN, SVM, Logistic Regression, LDA): One-hot encoding for nominal variables; ordinal encoding for ordered variables (e.g., speed bands)
-   **High-cardinality**: Pool `LGA` levels with \<1% of observations into `"Other"` to reduce dimensionality

### Step 7: Feature Scaling

-   **kNN and SVM** are distance-based and sensitive to feature scale → apply **z-score standardisation** to all numeric features within each training fold
-   **Tree-based models** are scale-invariant → no scaling needed

### Step 8: Class Imbalance Handling

The target `Degree of crash` is highly imbalanced (Fatal: 1.5%, Injury: 66.4%, Non-casualty: 32.1%). Resampling will be applied **within the training fold only** to prevent leakage (Chawla et al., 2002):

1.  **SMOTE** (`themis::step_smote`): Synthesise minority-class samples, especially for Fatal
2.  **Down-sampling**: Undersample the majority class as a simpler alternative
3.  Both strategies will be compared during model tuning to select the best approach

### Step 9: Train/Test Split & Cross-Validation

-   **Holdout**: 80/20 stratified split by `Degree of crash`
-   **Nested CV**: Outer 5-fold for generalisation error; inner 5-fold for hyperparameter tuning
-   All preprocessing steps will be embedded in a `tidymodels::recipe` pipeline so that transformations are fitted on training data only within each fold

### Pipeline Summary

```         
Raw Data (CRASH + TRAFFIC UNIT)
  │
  ├─ 1. Aggregate TRAFFIC UNIT → crash-level features
  ├─ 2. Left join on Crash ID
  ├─ 3. Remove identifiers, leakage vars, near-zero-variance cols
  ├─ 4. Feature engineering (binary flags, speed bands, temporal)
  ├─ 5. Outlier removal (geospatial bounds)
  ├─ 6. Train/Test split (80/20, stratified)
  │
  └─ Within each CV fold (training only):
       ├─ 7. Missing value handling
       ├─ 8. Categorical encoding (one-hot / factor)
       ├─ 9. Feature scaling (for kNN/SVM/LR)
       └─ 10. SMOTE / down-sampling for class imbalance
```

## 4.2 Model Plan: based on EDA, propose at least 5 classification models* suited to the data/problem with justification.

Based on the findings from EDA, we propose six classification models to predict **Degree of Crash** (a three-class outcome: Fatal, Injury, and Non-casualty). The EDA revealed that the target variable is extremely imbalanced—Fatal crashes account for a small proportion of the 92,189 observations—and that all predictor variables are categorical (e.g., `Weather`, `Natural lighting`, `Surface condition`, `Speed limit`, `Urbanisation`, `TU type group`). These characteristics directly inform the choice of models below. 

All models will be implemented in R using the `caret` package for a unified training and cross-validation interface.

### 1. Multinomial Logistic Regression (Baseline)
- **Justification:** Proposed as our interpretable baseline model. It produces interpretable coefficients (log-odds) for each predictor category, which are meaningful for road safety stakeholders (e.g., quantifying how much driving at night increases the log-odds of a fatal crash relative to daylight). All subsequent models will be benchmarked against this baseline.
- **Implementation:** Categorical predictors will be encoded as dummy variables. We will use `multinom()` from the `nnet` package, utilising the `class_weight` argument to account for class imbalance.

### 2. Decision Tree Classifier (CART)
- **Justification:** Chosen for its full interpretability and ability to capture non-linear interactions between categorical features (e.g., capturing the joint rule "high speed limit AND rural area $\rightarrow$ higher fatal risk" in a single branch).
- **Implementation:** Implemented using the `rpart` package. Tree depth will be tuned via cross-validation to prevent overfitting, and the resulting tree will be visualised using `rpart.plot` to communicate high-risk conditions to non-technical audiences.

### 3. Random Forest Classifier
- **Justification:** Extends the decision tree by training an ensemble of trees on bootstrap samples, significantly reducing variance. It provides variable importance scores to identify which predictors most strongly impact crash severity. 
- **Implementation:** With >90k observations, it is computationally feasible using the `randomForest` package. Class imbalance will be handled via the `classwt` argument in `randomForest()`.

### 4. Gradient Boosting Machine (XGBoost)
- **Justification:** Proposed as our primary high-performance model. By sequentially correcting misclassifications from prior trees, it is particularly effective at learning from the rare minority class (Fatal crashes) by exploiting subtle, high-value signals in the features.
- **Implementation:** Implemented using the `xgboost` package. Hyperparameters (`nrounds`, `max_depth`, `eta`) will be tuned via `caret::trainControl`. The `scale_pos_weight` parameter will be adjusted to address the class imbalance.

### 5. K-Nearest Neighbours (KNN)
- **Justification:** Included as a non-parametric, instance-based comparison model to test whether crash severity is locally consistent in feature space (i.e., whether crashes occurring under similar conditions tend to share the same outcome).
- **Implementation:** Implemented using the `class` package. Since all features are categorical, the Gower distance metric (via the `gower` package) will be used. The optimal value of $k$ will be selected via cross-validation.

### 6. Naive Bayes
- **Justification:** A natural fit for datasets where all features are categorical. It serves as a strong probabilistic benchmark: if more complex models cannot substantially outperform it, this suggests that the individual distributions of features (e.g., Weather, Speed limit) already carry most of the predictive information.
- **Implementation:** Implemented using the `e1071` package. Under the conditional independence assumption, it is extremely fast to train and provides probability estimates for each severity class.

## 4.3 Evaluation Plan: propose and justify performance metrics for final model 
evaluation.

Standard accuracy is an inadequate evaluation metric for this problem due to severe class imbalance. A classifier predicting *Injury* for every observation would achieve high accuracy while completely failing to identify any *Fatal* crash. The metrics below are selected to reflect both overall classifier performance and the disproportionate importance of correctly identifying fatal outcomes.

### Performance Metrics

- **Macro-Averaged F1 Score (Primary):** Computes the F1 score for each of the three classes independently and then averages them with equal weight. This ensures that performance on the rare *Fatal* class receives the same weight as the dominant *Injury* class. This will be our primary metric for ranking models.
- **Fatal Class Recall (Primary Constraint):** Measures the proportion of actual *Fatal* crashes correctly identified. A model with high Macro F1 but low Fatal recall will be deprioritised, as failing to flag potentially fatal conditions is the most consequential error.
- **Weighted F1 Score (Secondary):** Averages F1 per class weighted by each class's support. It provides a realistic picture of overall performance on the true class distribution.
- **Cohen's Kappa ($\kappa$) (Secondary):** Adjusts accuracy for the agreement expected by chance. Computed directly by `caret::confusionMatrix()`, providing a clear sanity check (values near 0 indicate performance no better than random guessing).
- **Confusion Matrix:** Evaluated via `caret::confusionMatrix()` and visualised as a heatmap using `ggplot2` to explicitly monitor misclassification directions (e.g., dangerous errors where Fatal crashes are predicted as Injury).

### Validation Strategy

- **Cross-Validation:** Stratified 5-fold cross-validation will be applied to the training data using `caret::trainControl(method = "cv", number = 5, classProbs = TRUE, savePredictions = TRUE)`. This ensures class proportions are preserved across all folds.
- **Temporal Split:** We will use a time-based holdout strategy: **2020–2023 data for training/CV** and **2024 data as the held-out test set**. This reflects real-world deployment conditions where historical data predicts future crashes.
- **Unbiased Estimation:** The final held-out test set (2024) will be evaluated *once only* after model selection and tuning are complete to obtain an unbiased estimate of generalisation performance.

## 4.4 Timeline and Project Milestones

The project timeline has been structured in alignment with the assignment 
deadlines to ensure clear and systematic progression of tasks from understanding 
the dataset up-to final model development and reporting. 

**1-7 April: Problem Framing and Data Acquisition**

Different datasets were evaluated and a final dataset was selected, along with 
finalization of the research question. The crash and Traffic Unit datasets (2020
-2024) were sourced from the NSW Government open data platform. Initial 
inspection was carried out to understand the dataset structure, key variables, 
and differences. 

**7-10 April: Data Integration and Initial Data Preparation**

The crash and traffic unit datasets were merged using Crash ID, resulting in a 
unified dataset containing 170,962 records at the vehicle level. Initial data 
preparation involved examining missing values, identifying inconsistencies 
introduced via the merging, and understanding different variable types. Full 
preprocessing and feature engineering will be conducted in the modelling phase. 

**10-14 April: Exploratory Data Analysis (EDA)**

Univariate and bivariate analyses were performed to explore the distribution and 
relationship between variables. This includes examining environmental factors 
(weather, lighting), temporal variables(time of day, day of week), and vehicle-
level characteristics (vehicle type, manoeuvre). Insights from EDA were used to 
inform feature selection and guide model planning.  

**14-17 April: Feature Planning and Model Design**

Based on EDA findings, relevant features were identified and transformations 
were planned, including the plan to encode categorical variables and aggregate 
vehicle-level information. A set of classification models were also selected for 
implementation in the next phase.

**17-19 April: Plan & EDA Report Finalization**

All findings were compiled into a structured R Markdown report and submitted by 
the deadline.

**6-18 May: Model Development and Refinement**

Classification models will be implemented, including baseline and advanced 
approaches. Preprocessing steps such as handling missing values, encoding 
categorical variables, and feature engineering will be carried out. Model 
refinement including hyperparameter tuning, will also be conducted to improve 
performance.

**18-24 May: Model Evaluation and Final Report Preparation**

Models will be evaluated using appropriate performance metrics, considering 
class imbalance in crash severity. The final model will be selected and 
justified. Results, insights, and interpretations will be consolidated into a 
final report.

**25-31 May: Final Checks and Contribution Review**

Final refinements will be made to ensure clarity and consistency of the report. 
Team contributions will be reviewed and documented.

# References

List all the references used throughout the report using APA7 Style.

# AI Use Statment
If applicable,

I acknowledge the use of [AI tool + version, publisher and URL] to generate [summary of what was generated].
